White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-12-17 papers/wdmerger_I/paper.tex
dx -> delta x
Shorten abstract to fit within 250 word ApJ limit

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possible progenitor that has 
merited recent theoretical attention is the white dwarf merger scenario.
This has the potential to naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been relatively few fully self-consistent simulations 
of merging binary white dwarf systems using mesh-based hydrodynamics. 
This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
adaptive mesh refinement. In this paper we describe our numerical 
methodology and discuss our implementation in the compressible hydrodynamics code \castro.
For merging white dwarf simulations we use \castro's capability to solve the Euler equations, and
the Poisson equation for self-gravity, and couple the gravitational force to the hydrodynamics.
Standard techniques for coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system for our problem, 
but recent advances in the literature allow progress on this 
issue and we discuss our implementation here. We present a
set of test problems demonstrating the extent to which our software sufficiently
models a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, using the initialization procedure described below 
in \autoref{sec:implementation}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by \autoref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in \autoref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. In \autoref{sec:kepler} we 
show that there are no gains to be had by increasing the accuracy further. At 
very high orders ($l \gtrsim 18$) the approximation breaks down, as seen in \autoref{fig:bc_comparison}. 
This is a result of the ambient material on the grid. At each boundary point we 
assume that all of the mass on the grid is contained within a sphere whose radius is 
the distance from that boundary point to the center of the domain. This does not hold 
for boundary points in the centers of domain faces, because of the material in the 
domain corners. This can be fixed by using multiple mass shells at diferent radii, but the error 
is negligible in practice for the values of $l_{\text{max}}$ that we use.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

>>>>

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}

The Type Ia supernova progenitor problem is one of the most perplexing and
exciting problems in astrophysics, requiring detailed numerical modeling to
complement observations of these explosions. One possible progenitor that has
merited recent theoretical attention is the white dwarf merger scenario, which
has the potential to naturally explain many of the observed characteristics of
Type Ia supernovae. To date there have been relatively few self-consistent simulations
of merging white dwarf systems using mesh-based hydrodynamics. This is the first
paper in a series describing simulations of these systems using a hydrodynamics
code with adaptive mesh refinement. In this paper we describe our numerical
methodology and discuss our implementation in the compressible hydrodynamics
code CASTRO, which solves the Euler equations, and the Poisson equation for
self-gravity, and couples the gravitational and rotation forces to the hydrodynamics.
Standard techniques for coupling gravitation and rotation forces to the hydrodynamics
do not adequately conserve the total energy of the system for our problem, but
recent advances in the literature allow progress and we discuss our implementation
here. We present a set of test problems demonstrating the extent to which our
software sufficiently models a system where large amounts of mass are advected
on the computational domain over long timescales. Future papers in this series
will describe our treatment of the initial conditions of these systems and will
examine the early phases of the merger to determine its viability for triggering
a thermonuclear detonation.

@@

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, using the initialization procedure described below 
in \autoref{sec:implementation}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by \autoref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} \Delta x\, \Delta y\, \Delta z\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in \autoref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. In \autoref{sec:kepler} we 
show that there are no gains to be had by increasing the accuracy further. At 
very high orders ($l \gtrsim 18$) the approximation breaks down, as seen in \autoref{fig:bc_comparison}. 
This is a result of the ambient material on the grid. At each boundary point we 
assume that all of the mass on the grid is contained within a sphere whose radius is 
the distance from that boundary point to the center of the domain. This does not hold 
for boundary points in the centers of domain faces, because of the material in the 
domain corners. This can be fixed by using multiple mass shells at diferent radii, but the error 
is negligible in practice for the values of $l_{\text{max}}$ that we use.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}
